Investigating Determinism of Game Engines for Autonomous Vehicle Verification

### 2020-04-20 paper/cav_determinisim_main.tex
New branch with AG's updates

% Requirements of simulation based verification SBV
The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses. 
%
Furthermore, they are easy to setup and run; they also are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}. 
%
For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models. Carla currently offers various basic types of cameras and ray-case based lidar; extensions are still under development. 
%
Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required.

@@

Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.  %However, when these engines are used for AV development then deterministic behaviour is required. 
This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
%
% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations. Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.

@@

The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
The shorter this time step is, the smoother the interpretation of the physical dynamics will be. %processing time it takes to catch up to game clock. % and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}. 
%
Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
%
To use a fixed physics time step the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one render tick (one frame of real world time). Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}. 
%
Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware. Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/}}. 
%
If the engine tries to render between physics updates, \textit{residual lag} can occur and extrapolation between frames needs to be performed to smooth transition between scenes. This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited?}\todo[inline]{Why is this part of the figure? It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.}
%
% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.\todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?}

@@

\begin{figure*}[b]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
    \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation environment.}

@@

\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance. This is particularly important for perception-stack testing (the sensors need to get input that is as close to the real input as possible, ideally without extrapolation or lag) and visual inspection of accidents or near misses (same is true for a human observer). For instance, see below, but I'm not sure whether this is the right place.}
In addition, providing game players with a responsive real-time experience is often achieved at the cost of realism.\todo[inline]{Do we have a ref for this, Unity vs UE4?}
%
In contrast, realism is essential in AV simulation, especially for perception stack testing, but also for visual inspection. If necessary, it is acceptable to achieve the required level of realism at the cost of the real-time user experience.

@@

\noindent\underline{\textit{Concurrency:}}
% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism. Consider also parallelism, e.g. in GPUs. Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism. Say which parts of a game engine could reasonably be expected to run concurrently.}
Another issue to consider is concurrency, where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program. This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}. A game engine may use such techniques to improve performance.\todo[inline]{how?}
% KIE: We only list the potential sources of non-det, I therefore removed this. 
% although monitoring and detecting such events falls outside of the scope for this work.

@@

% ***************************************************
%  METHOD
% ***************************************************
\section{Methodology} \label{s:methodology}
In this section a method for determining the variance of a simulation environment, termed \textit{simulation variance} is presented. We have identified a number of experimental parameters that may influence the simulation variance of a game engine, see Fig.~\ref{method_diagram}. These need to be considered at different stages in the process.  
%
The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine. These may affect different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
\todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''.} 
%
To replicate and control the high computational loads that may be experienced in high performance simulations, software that artificially utilises resources is executed alongside the simulation on both central and graphical processing units. 
\todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW.}
% The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system. The aim of the experiment is to determine if the simulation is deterministic. This is achieved by monitoring actors in repeated tests and examining the deviation in path trajectories. Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.

\subsection{Setting a Determinism  Threshold}
We use the term \textit{determinism threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits. At some low analogue voltage the signal is interpreted as binary $0$, this may be below 0.5V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above 3.5V.

The level of variance acceptable for the simulation must first be set. It serves as a threshold for deterministic execution. This level depends on the objectives of the simulation and the granularity at which the simulation environment operates. 
%
\todo{KIE HERE}
In our case study, presented in Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required. This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment but a larger value may be acceptable for non-urban environments, e.g. motorway or highway. Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g. $1\mu m$. Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios. Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.

@@

% We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians). Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g. A* search algorithm.

\subsection{System Resources}
To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources. Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will  contribute towards increased \textit{simulation variance}. CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program that can be used spawn workers on any number of cores or virtual threads on a system. This can be used to artificially increased the load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}. For GPU utilisation \texttt{gpu-burn} was employed using the fur test at different resolutions and multiple instances can be used to generate high utilisation levels~\cite{GPU_stress}.

@@

\subsection{Internal settings}
Where possible a fixed $dt$ should be used to give the most reliable physics calculations. If using Carla a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}} and a good description of the interplay between physics and render clock is given here~\cite{JohnAustinUnity}. In Carla, further settings can be applied to allow synchronous communication to external controllers to ensure no sensor data are passed out of order to the simulator~\cite{carla_sim_config}.

@@

\subsection{Output and Analysis}
The output for each experiment can be summarised as the \textit{maximum variance} of any actor over the $n-$repeated tests. The maximum variance is chosen to indicate the worst case rather than an average which could hide outlier data points when a high number of repeated tests are performed. This value can be analysed against the different scenarios or settings that have been identified as potential sources of simulation variance. These values can then be assessed against the user-defined \textit{deterministic threshold} to determine if the simulation is deterministic or not.

>>>>

% Requirements of simulation based verification SBV
The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes suitable for perception stack testing. 
%
Furthermore, they are easy to setup and run; they also are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}. 
%
For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models. Carla currently offers various basic types of cameras and ray-case based lidar; extensions are still under development. 
%
Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required.

@@

Investigating the determinism of game engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable execution. %However, when these engines are used for AV development then deterministic behaviour is required. 
This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
%
% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations. Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.

@@

The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
The shorter this time step is, the smoother the interpretation of the physical dynamics will be. %processing time it takes to catch up to game clock. % and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}. 
%
Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
%
To use a fixed physics time step the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one render tick (one frame of real world time). Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}. 
%
Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware. Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/}}. 
%
If the engine tries to render between physics updates, \textit{residual lag} can occur and extrapolation between frames needs to be performed to smooth transition between scenes. This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited?}\todo[inline]{Why is this part of the figure? It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.}
%
% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.

@@

\begin{figure*}[b]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v2.pdf}
    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a game engine for usage in AV simulations.}

@@

@@

\noindent\underline{\textit{Concurrency:}}
\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism. Consider also parallelism, e.g. in GPUs. Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism. Say which parts of a game engine could reasonably be expected to run concurrently.}
Another issue to consider is concurrency, where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program. This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}. A game engine may use such techniques to improve performance.\todo[inline]{how?}
% KIE: We only list the potential sources of non-det, I therefore removed this. 
% although monitoring and detecting such events falls outside of the scope for this work.

@@

% ***************************************************
%  METHOD
% ***************************************************
\section{Methodology} \label{s:methodology}
In this section a method for determining \textit{simulation variance} of an end-user system is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}. The method includes scenarios that exercise different actor types and functional callbacks of the game engine that may tax different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks. To replicate high computational loads that may be experienced in high realism simulations, software that artificially utilises resources is used at runtime for both central and graphical processing units. 
% The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system. The aim of the experiment is to determine if the simulation is deterministic. This is achieved by monitoring actors in repeated tests and examining the deviation in path trajectories. Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.

% \subsection{Deterministic Threshold}
% We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. This threshold is analogous to the analogue signal level associated with a binary state in a logical processor. At some low analogue voltage the signal is interpreted as binary $0$ which may be below 0.5V and is considered binary $1$ when the same signal is above 3.5v.

% The user must specify the level of variance acceptable for their application which is used as a threshold for deterministic execution. In our case study, Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required. This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment but a larger value may be acceptable for non-urban environments, e.g. motorway or highway. Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g. $1\mu m$. Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios. Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.

@@

% We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians). Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g. A* search algorithm. 
\subsection{Execution} 
\subsubsection{System Resources}
To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources. Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will  contribute towards increased \textit{simulation variance}. CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program that can be used spawn workers on any number of cores or virtual threads on a system. This can be used to artificially increased the load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}. For GPU utilisation \texttt{gpu-burn} was employed using the fur test at different resolutions and multiple instances can be used to generate high utilisation levels~\cite{GPU_stress}.

@@

\subsubsection{Internal settings}
Where possible a fixed $dt$ should be used to give the most reliable physics calculations. If using Carla a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}} and a good description of the interplay between physics and render clock is given here~\cite{JohnAustinUnity}. In Carla, further settings can be applied to allow synchronous communication to external controllers to ensure no sensor data are passed out of order to the simulator~\cite{carla_sim_config}.

@@

\subsection{Output and Analysis}
% \subsection{Deterministic Threshold}
We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. This threshold is analogous to the analogue signal level associated with a binary state in a logical processor. At some low analogue voltage the signal is interpreted as binary $0$ which may be below 0.5V and is considered binary $1$ when the same signal is above 3.5v.

The user must specify the level of variance acceptable for their application which is used as a threshold for deterministic execution. In our case study, Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required. This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment but a larger value may be acceptable for non-urban environments, e.g. motorway or highway. Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g. $1\mu m$. Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios. Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.

The output for each experiment can be summarised as the \textit{maximum variance} of any actor over the $n-$repeated tests. The maximum variance is chosen to indicate the worst case rather than an average which could hide outlier data points when a high number of repeated tests are performed. This value can be analysed against the different scenarios or settings that have been identified as potential sources of simulation variance. These values can then be assessed against the user-defined \textit{deterministic threshold} to determine if the simulation is deterministic or not.
